### 2022-02-17 README.Rmd
introduce bayes

A successful prediction should lend more credibility to a theory than a postdiction, all else equal.
Much has been theorized on this important distinction between exploration and confirmation, discovery and justification, hypothesis generating and hypothesis testing and so forth (see XXXX).
To honor this distinction, many psychological studies are preregistered to produce results that are considered confirmatory.
Rigorous application of preregistration prevents researcher from reporting a simple confirmatory story for a set of results that were produced by an arduous process of try and error.
The scientific community, widely agrees that actively concealing such "trying for results" is misconduct.
But what exactly is such concealment cheating the scientific community of?
We believe that a rigorous answer to this question enables us to clear some common misunderstandings that lead researchers to deem preregistration as unfit for their purposes.
These misunderstandings are interesting from a theoretical perspective but also lead to decreased efficiency of the scientific process if researcher  misguidedly decide that preregistration is not applicable for their intended study.

One reason why researchers are hesitant to preregister is that they feel too limited by preregistration.
Even researchers who have been persuaded by its advantages sometimes apply strategies to skirt the self imposed restrictions.
One strategy is to write a loose preregistration to begin with, another is to simply deviate from the preregistration afterward.
To complicate matters further, both strategies may be used with compelling scientific reasoning or with the self-serving intent of generating good looking results no matter the nature of the phenomenon under study.
To summarize, the conundrum is that current conceptualization of preregistration deem a study "properly" preregistrated even if it contains a recipe for p-hacking before data collection but "improper" if the researcher did technically deviate afterwards.
This, of course, goes against common researchers intuition about what is considered exploratory and confirmatory.
We point these issues out as symptoms of a misunderstanding.#change
We do want not to discredit preregistration as a useful practice for advancing science, but challenge the common view that preregistration somehow cleanly separates confirmatory from exploratory.

Deciding if a study is confirmatory or exploratory is a judgment made by the researcher.
The aforementioned issues arise by delegating this judgment to a simple decision rule, that is, if something is preregistered it is confirmatory and if it is not preregistered it is exploratory.
Preregistering an analysis that is inherently exploratory (like testing dozens of relations) does not make it confirmatory nor will a careful confirmatory study become exploratory if researcher deviate from the preregistration in minor details.
To uphold the simple decision rule, it is routinely suggested to only preregister when a study is expected to have a low type I error rate by design and will run/be analysed without changes no matter what.
Under these conditions the decision rule holds, but such restricted use makes preregistration a niche solution not able to match the greater problem of replicability in psychology and elsewhere.

@@

Let us start by defining what we mean by expected epistemic value.
If researchers plan to conduct a study they usually hope it will change their assessment of the verisimilitude (truthlikeness) of some theory.
In other words, they hope to learn something from conducting the study.
The amount of knowledge researchers gain from a particular study in relation to a specific theory is what we call epistemic value.
While researcher can not know what exactly they will learn from a study, they can form an expectation.
This expectation is what we call expected epistemic value.
We assume three things about this process of estimation and how it relates to choosing a study to conduct.

@@

The first assumption leads to our adoption of a Bayesian framework.
Our rational is as follows.
Researchers that decide to conduct a study are akin to choosing a study to bet on.
They have to "place the bet" by conducting the study therefore invest resources and stand to gain epistemic value with some probability.
This conceptualization of choosing a study to conduct as a betting problem, lets us apply a "Dutch Book" argument.
This argument, states that any better must follow the axioms of probability if they want to avoid being "irrational", i.e., accept bets with sure losses.
Fully developing a Dutch book argument for this problem requires a careful consideration of what kind of studies to include as possible bets, defining a conversion rate from the stakes to the reward, and modeling what liberties researchers have in choosing a study.
Without deliberating these concepts further, we find it persuasive that researchers should not violate the axioms of probability if they have some expectation about what they stand to gain with some likelihood from conducting a study.
The axioms of probability are all that is required to derive the famous Bayes formula, on which we will heavily rely for our arguments.
Please note that our decision to adopt a Bayesian philosophy of science does not imply anything about the statistical methods researcher use.
In fact, this conceptualization is purposefully reductionistic to be compatible with a wide range of philosophies of science and statistical methods researchers might subscribe to.
Our conclusions can be applied to bayesian and frequentist methods alike.
Of course, frequentist methods are traditionally viewed from a poperian philosophy of science, but our conclusions are compatible with it and notable later advancements like the error statistical view.
However, those philosophies of science require a much richer set of assumptions that are not quite necessary for the points we want to make.

@@

Let us make some simplifying assumptions and define notation.
We restrict ourselfs to consider evidence of a binary nature (either exists or does not), since continues evidence would led to some quite involved derivations.
We denote the probability of a theory before observing evidence as $P(H)$ and its complement as $P(¬H) = 1 - P(H)$.
The probability of observing evidence under some hypothesis is $P(E|H)$.
The total probability of evidence, $P(E)$, is given by $P(H)P(E|H) + P(¬H)P(E|¬H)$.
From these probabilities we can calculate the probability of the hypothesis after observing the evidence via:

@@

From the three probabilities involved in the calculation $P(E|¬H)$ is the one our argument rests on.

>>>>

A successful prediction should lend more credibility to a theory than a postdiction, all else equal.
Much has been theorized on this important distinction between exploration and confirmation, discovery and justification, hypothesis generating and hypothesis testing and so forth (see XXXX).
To honor this distinction, many psychological studies are preregistered to produce results that are considered confirmatory.
Rigorous application of preregistration prevents researcher from reporting a simple confirmatory story for a set of results that were produced by an arduous process of try and error.
The scientific community, widely agrees that actively concealing such "trying for results" is misconduct.
But how is such concealment hurting the scientific process?
We believe that a rigorous answer to this question enables us to clear some common misunderstandings that lead researchers to deem preregistration as unfit for their purposes.
These misunderstandings are interesting from a theoretical perspective but also lead to decreased efficiency of the scientific process if researcher misguidedly decide that preregistration is not applicable for their intended study.

One reason why researchers are hesitant to preregister is that they feel too limited by preregistration.
Even researchers who have been persuaded by its advantages sometimes apply strategies to skirt the self imposed restrictions.
One strategy is to write a loose preregistration to begin with, another is to simply deviate from the preregistration afterward.
To complicate matters further, both strategies may be used with compelling scientific reasoning or with the self-serving intent of generating good looking results no matter the nature of the phenomenon under study.
To summarize, the conundrum is that current conceptualization of preregistration deem a study "properly" preregistrated even if it contains a recipe for p-hacking before data collection but "improper" if the researcher did technically deviate afterwards.
This, of course, goes against common researchers intuition about what is considered exploratory and confirmatory.
We take these issues as signals of a misunderstanding about how preregistration contributes to the scientific progress.
In particular, we want to challenge the common view that preregistration somehow cleanly separates confirmatory from exploratory.

Deciding if a study is confirmatory or exploratory is a judgment made by the researcher.
The aforementioned issues arise by delegating this judgment to a simple decision rule, that is, if something is preregistered it is confirmatory and if it is not preregistered it is exploratory.
This rule is often assumed implicitly as a rule of thumb, other times it is stated as the explicit goal of preregistration, but always leads to confusion when researchers try to apply it outside specific bounds.
Preregistering an analysis that is inherently exploratory (like testing dozens of relations) does not make it confirmatory nor will a careful confirmatory study become exploratory if researcher deviate from the preregistration in minor details.
To uphold the simplilicity of the decision rule, it is routinely suggested to only preregister when a study is expected to have a low type I error rate by design and will run/be analysed without changes no matter what.
Under these conditions the rule holds, but such restricted use makes preregistration a niche solution not able to match the greater problem of replicability in psychology and elsewhere.

@@

Let us start by defining what we mean by expected epistemic value.
If researchers plan to conduct a study they usually hope it will change their assessment of the verisimilitude (truthlikeness) of some theory.
In other words, they hope to learn something from conducting the study.
The amount of knowledge researchers gain from a particular study in relation to the versimilitude of a specific theory is what we call epistemic value.
While researcher can not know what exactly they will learn from a study, they can form an expectation.
This expectation is what we call expected epistemic value.
We assume three things about this process of estimation and how it relates to choosing a study to conduct.

@@

The first assumption leads to our adoption of a Bayesian framework.
Our rational is as follows.
Researchers that decide to conduct a study are akin to choosing a study to bet on.
They have to "place the bet" by conducting the study therefore invest resources and stand to gain epistemic value with some probability.
This conceptualization of choosing a study to conduct as a betting problem, lets us apply a "Dutch Book" argument.
This argument, states that any better must follow the axioms of probability if they want to avoid being "irrational", i.e., accept bets with sure losses.
Fully developing a Dutch book argument for this problem requires a careful consideration of what kind of studies to include as possible bets, defining a conversion rate from the stakes to the reward, and modeling what liberties researchers have in choosing a study.
Without deliberating these concepts further, we find it persuasive that researchers should not violate the axioms of probability if they have some expectation about what they stand to gain with some likelihood from conducting a study.
The axioms of probability are all that is required to derive the famous Bayes formula, on which we will heavily rely for our arguments.
Please note that our decision to adopt this aspect of Bayesian philosophy of science does not imply anything about the statistical methods researcher use.
In fact, this conceptualization is purposefully reductionistic to be compatible with a wide range of philosophies of science and statistical methods researchers might subscribe to.
Our conclusions can be applied to bayesian and frequentist methods alike.
Of course, frequentist methods are traditionally viewed from a poperian philosophy of science, but our conclusions are compatible with this view and and notable later advancements like the error statistical view.
However, those philosophies of science require a much richer set of assumptions that are not quite necessary for the points we want to make about the value of preregistration.

@@

Let us make some simplifying assumptions and define notation.
We restrict ourselfs to consider evidence of a binary nature (either exists or does not), since continues evidence would led to some quite involved derivations.
We denote the probability of a theory before observing evidence as $P(H)$ and its complement as $P(¬H) = 1 - P(H)$.
The probability of observing evidence under some hypothesis is $P(E|H)$.
We can calculate the probability of the hypothesis after observing the evidence with help from the Bayes formula:

$$
P(H|E) = \frac{P(H)P(E|H)}{P(E)}
$$

There we can replace the total probability of evidence, $P(E)$ with $P(H)P(E|H) + P(¬H)P(E|¬H)$ and arrive at:

@@

```{r, include=FALSE}
bayes <- function(h, eh, enh)(h * eh)/((h * eh) + ((1 - h) * enh))
```

From the three probabilities involved in the calculation, $P(E|¬H)$ is the one our argument rests on.
It tells use how likely it is that we find evidence, when in fact the theory is not true.
How $P(E|¬H)$ relates to $P(H|E)$ is of great relevance, since $P(H|E)$ is often used directly or indirectly as a measure of corroboration of a hypothesis.
In carnaps tradition, in its direct use, it is called corroboration as firmness, in its relation to $P(H)$ it is called increase in firmness.
We refrain from discussing specific measures of corroboration, since there is no measure that show universally better properties than others.
However, it is generally expected that any measure of corroboration increases monotonically with an increase in $P(H|E)$.

In short, we want to increase $P(H|E)$.
Increases in $P(H|E)$ are associated with increases epistemic value, of which we want to maximize the expectation.
So how can we increase $P(H|E)$?
The Bayes formula gives us three options to investigate, namely $P(H)$, $P(E|H)$ and $P(E|¬H)$.
Before we continue, let us note that $P(E|H)$ and $P(E|¬H)$ are not complements (they do not necessarily sum to 1).
The first option leads us to the unsurprising conclusion that higher $P(H)$ leads to higher $P(H|E)$, but the a priori probability of a hypothesis is nothing our study design can change.
The second option is similar commonsense, that is an increase $P(E|H)$ leads to higher $P(H|E)$.
Researcher therefore should make sure that they design their studies in a way that in case their hypothesis is true, they can actually find evidence for it.
While this truism is strongly related to study design, it is unrelated to preregistration.
The third option, while mathematically equally straightforward, is not quite as intuitive.
It concerns $P(E|¬H)$, therefore the probability of observing evidence for the theory, when in fact the theory is wrong.
Before we discuss what this probability means and how it relates to preregistration, let us show how it impacts $P(H|E)$.
First, increasing $P(E|¬H)$ does decrease $P(H|E)$.

## How statistical method contribute to $P(E|¬H)$

For the overly simplistic case that the hypothesis is a statistical nill-hypothesis, $P(E|¬H)$ becomes the type I error rate.
Researchers who choose a smaller type I error rate, can therefore be surer of their results.
However, the research hypothesis is seldomly equal to the statistical null hypothesis.
$P(E|¬H)$ encompasses factors also outside of the statistical realm, like the studies design.
Statistical methods have the great advantage that we have a large toolbox of assessing and controlling their contribution to $P(E|¬H)$.
Examples of our ability to exert control are the setting of type I error rate, the use of corrected fit measures (i.e., adjusted R²) or cross validation in machine learning.
These are all tools that help us to factor in the tendency of statistical methods to show results even when in fact their are none and are therefore closely related to the probability of observing evidence when in fact there is none.
